\documentclass[11pt,oneside]{book}
\usepackage{uwnotes}

\title{The Mathematics of the Music Microscope}
\author{Harry Geng}
\date{}

\begin{document}

% ================= title page =================
\begin{titlepage}
\centering
{\Huge\bfseries The Mathematics of the\\[4pt] Music Microscope\par}
\vspace{1.2cm}
{\Large Course Notes for the Audio-Visualizer Project\par}
\vspace{2.2cm}
{\large Harry Geng\par}
\vfill
{\small Living document --- grows with the codebase.\\
Chapters marked \emph{(outline)} are planned but not yet written.\par}
\vspace{0.8cm}
{\small \today\par}
\end{titlepage}

\frontmatter
\tableofcontents

% ================= preface =================
\chapter{Preface}
These notes document, with proofs where they are illuminating, the mathematics
running inside the Audio-Visualizer project: a system that ingests songs,
separates them into stems with neural networks, extracts musical structure,
embeds every few seconds of audio into learned vector spaces, and renders the
result as interactive visualizations.

The material spans four traditions that rarely share a book: classical signal
processing (Part~\ref{part:signals}), modern deep learning
(Part~\ref{part:learning}), the geometry of high-dimensional data
(Part~\ref{part:geometry}), and computer graphics
(Part~\ref{part:rendering}). The unifying thread is that \emph{every object
defined here actually runs}: each chapter ends where the repository begins,
and the ``In the code'' boxes point at the specific file, function, and
parameter values where a definition earns its keep. When a theorem's
hypothesis is violated in practice (they often are), the notes say so and
describe what breaks.

Exercises range from short proofs to experiments you can run against your own
music library. None are busywork; several reproduce bugs that were actually
encountered while building the system.

\section*{How to read this}
Chapter~\ref{ch:sound} and Chapter~\ref{ch:timbre} are complete and set the
standard for the rest. The remaining chapters exist as detailed outlines and
will be filled in as the corresponding subsystems stabilize. Prerequisites are
first-year linear algebra and calculus; everything else is developed as
needed.

\section*{Notation}
\begin{center}
\begin{tabular}{ll}
\toprule
$x[n]$ & a discrete-time signal (square brackets: integer index) \\
$f_s$ & sampling rate in Hz \\
$X[k]$ & the DFT of $x[n]$; $k$ indexes frequency bins \\
$\norm{x}$ & Euclidean norm; $\norm{x}_F$ the Frobenius norm \\
$\ip{x,y}$ & inner product $x^\T y$ \\
$S^{d-1}$ & the unit sphere $\{x \in \R^d : \norm{x} = 1\}$ \\
$A^\T$ & matrix transpose \\
$\mathbf{1}[\cdot]$ & indicator function \\
$\log$ & natural logarithm unless a base is written \\
\bottomrule
\end{tabular}
\end{center}

\mainmatter

% ================= PART I =================
\part{Signals}\label{part:signals}
\input{chapters/01-sound-to-vectors}
\input{chapters/02-filters-beats}

% ================= PART II =================
\part{Learning}\label{part:learning}
\input{chapters/03-attention}
\input{chapters/04-selfsupervised}
\input{chapters/05-separation}
\input{chapters/06-contrastive}

% ================= PART III =================
\part{Geometry of Timbre}\label{part:geometry}
\input{chapters/07-timbre-maps}
\input{chapters/08-similarity}

% ================= PART IV =================
\part{Rendering}\label{part:rendering}
\input{chapters/09-shader-math}

% ================= appendix =================
\appendix
\input{chapters/a-equal-temperament}

\end{document}
