\chapter{Attention and Transformers \emph{(outline)}}\label{ch:attention}
\chapterlede{The workhorse architecture behind every neural network in this
project: MERT, CLAP's text tower, Demucs' bottleneck, and BS-RoFormer are all
transformers. This chapter builds the mechanism once so later chapters can
use it freely.}

\begin{incode}
Status: outline. The definitions here are prerequisites for
Chapters~\ref{ch:selfsup}--\ref{ch:contrastive}.
\end{incode}

Planned contents:
\begin{itemize}
  \item \textbf{Scaled dot-product attention.}
  $\operatorname{Attn}(Q,K,V) = \softmax(QK^\T/\sqrt{d_k})\,V$; attention as
  a data-dependent convex combination; the $\sqrt{d_k}$ scaling justified via
  the variance of dot products of random vectors (with proof).
  \item \textbf{Multi-head attention} as subspace projections; parameter
  counting; why heads specialize.
  \item \textbf{Positional information.} Sinusoidal encodings; relative
  position; \emph{rotary embeddings} (RoPE) --- rotation matrices acting on
  pairs of coordinates, and the theorem that RoPE makes attention scores a
  function of relative offset only (this is the ``Ro'' in BS-RoFormer).
  \item \textbf{The full block.} Residual streams, LayerNorm (with the
  gradient argument for pre-norm), feed-forward expansion; $O(T^2)$ cost in
  sequence length and what that means for $75$\,fps audio frames.
  \item \textbf{Softmax and temperature.} Properties used throughout:
  invariance to additive shifts, the Jacobian, low/high temperature limits.
\end{itemize}
