\chapter{Maps of Timbre Space: PCA and UMAP}\label{ch:timbre}
\chapterlede{The galaxy view draws every moment of every song as a star, and
similar-sounding moments land near each other. This chapter contains the
complete mathematics of that map: from $768$-dimensional learned embeddings,
through an optimal linear compression (PCA), to a nonlinear graph embedding
(UMAP) whose axes provably mean nothing --- and whose neighborhoods mean
everything.}

\section{The Data: Moments on a Sphere}

Chapter~\ref{ch:selfsup} describes how MERT turns each $\sim$4-second moment
of audio into a vector $e \in \R^{768}$ by mean-pooling transformer states.
Here we take those vectors as given. The pipeline first normalizes:
\[
x \;=\; \frac{e}{\norm{e}} \;\in\; S^{767},
\]
placing every moment on the unit sphere. This makes similarity purely
\emph{angular}:

\begin{definition}[Cosine similarity and distance]\label{def:cosine}
For nonzero $u, v \in \R^d$,
\[
\operatorname{sim}(u,v) = \frac{\ip{u,v}}{\norm{u}\norm{v}},
\qquad
d_{\cos}(u,v) = 1 - \operatorname{sim}(u,v) \;\in\; [0, 2].
\]
For unit vectors, $\operatorname{sim}(u,v) = \ip{u,v}$ and
$\norm{u - v}^2 = 2\, d_{\cos}(u,v)$: cosine distance and squared Euclidean
distance are the same thing on the sphere, up to a factor of $2$.
\end{definition}

\begin{remark}
$d_{\cos}$ is not a metric (it fails the triangle inequality in general), but
the angle $\arccos(\operatorname{sim})$ is. In practice every algorithm below
only ever \emph{ranks} distances, for which the distinction is harmless.
\end{remark}

\begin{incode}
\texttt{compute\_galaxy.py} stacks the \texttt{emb\_mix} arrays of every song
into $X \in \R^{n \times 768}$ and normalizes rows
(\texttt{X /= np.linalg.norm(X, axis=1, keepdims=True)}). With the current
local library, $n = 664$; the PC batch library will push $n$ into the tens of
thousands, which is where every guarantee in this chapter gets sharper.
\end{incode}

\section{Principal Component Analysis}

The first reduction $768 \to 64$ is linear, classical, and optimal in two
senses that turn out to be the same sense.

Let $X \in \R^{n \times d}$ have centered columns (subtract the mean row
$\bar{x}$ first), and define the sample covariance
\[
C = \frac{1}{n-1} X^\T X \;\in\; \R^{d \times d}.
\]
$C$ is symmetric positive semidefinite, so by the spectral theorem it has an
orthonormal eigenbasis $v_1, \dots, v_d$ with eigenvalues
$\lambda_1 \ge \lambda_2 \ge \cdots \ge \lambda_d \ge 0$.

\begin{theorem}[Variance maximization]\label{thm:pca}
The direction of maximal projected variance,
\[
\argmax_{\norm{u} = 1} \Var(Xu) = \argmax_{\norm{u} = 1} u^\T C u,
\]
is $v_1$, and the maximum equals $\lambda_1$. More generally, the
$k$-dimensional subspace maximizing total projected variance is
$\operatorname{span}(v_1, \dots, v_k)$, capturing variance
$\sum_{i=1}^{k} \lambda_i$.
\end{theorem}

\begin{proof}
Expand $u$ in the eigenbasis: $u = \sum_i \alpha_i v_i$ with
$\sum_i \alpha_i^2 = 1$. Then
\[
u^\T C u = \sum_{i} \lambda_i \alpha_i^2
\;\le\; \lambda_1 \sum_i \alpha_i^2 = \lambda_1,
\]
with equality iff $\alpha_i = 0$ whenever $\lambda_i < \lambda_1$. For the
subspace statement, apply the same argument inductively on the orthogonal
complement of the span chosen so far (the Courant--Fischer characterization).
\end{proof}

\begin{theorem}[Best low-rank approximation, Eckart--Young]\label{thm:eckart}
Write the (thin) singular value decomposition $X = U \Sigma V^\T$ with
singular values $\sigma_1 \ge \cdots \ge \sigma_d$. Among all matrices $M$ of
rank at most $k$,
\[
\norm{X - M}_F^2 \;\ge\; \sum_{i > k} \sigma_i^2,
\]
with equality for $M = U \Sigma_k V^\T$ ($\Sigma_k$ keeps the top $k$
singular values). Equivalently: projecting the rows of $X$ onto
$\operatorname{span}(v_1,\dots,v_k)$ is the rank-$k$ linear map minimizing
total squared reconstruction error.
\end{theorem}

\begin{remark}
Theorems~\ref{thm:pca} and \ref{thm:eckart} are the same fact seen from two
sides, connected by $C = \frac{1}{n-1} V \Sigma^2 V^\T$: the covariance
eigenvectors are the right singular vectors, and
$\lambda_i = \sigma_i^2/(n-1)$. ``Keep the most variance'' and ``lose the
least reconstruction'' pick the same subspace.
\end{remark}

\begin{incode}
\texttt{PCA(n\_components=64).fit\_transform(X)} in
\texttt{compute\_galaxy.py}. Two engineering motives, neither about
visualization per se: (i) UMAP's neighbor search runs $\sim$12$\times$
faster in $64$ dimensions than in $768$; (ii) the trailing dimensions of
mean-pooled transformer embeddings are mostly noise, and
Theorem~\ref{thm:eckart} says truncation is the \emph{least} destructive way
to drop them. Note that with $n = 664 < 768$ the data matrix cannot have rank
above $663$ anyway; PCA here is denoising, not compression.
\end{incode}

\section{UMAP I: The Graph}

UMAP\footnote{McInnes, Healy, Melville, \emph{UMAP: Uniform Manifold
Approximation and Projection for Dimension Reduction} (2018). These notes
present the algorithm as implemented, with the category-theoretic framing
omitted.} operates in two acts. Act one converts the point cloud
$\{x_1, \dots, x_n\} \subset \R^{64}$ into a weighted graph encoding local
structure; act two lays that graph out in the plane. The high-dimensional
coordinates never appear again after act one --- everything the map knows is
in the graph.

\begin{definition}[$k$-NN graph]\label{def:knn}
Fix $k \in \N$ (here $k = 15$) and a metric $d$ (here $d_{\cos}$). For each
$i$ let $N(i)$ be the set of its $k$ nearest neighbors. The directed graph
has an edge $i \to j$ for each $j \in N(i)$.
\end{definition}

Exact $k$-NN costs $\Theta(n^2 d)$; the implementation uses NN-descent, a
randomized refinement scheme, for near-linear approximate search.

The heart of UMAP is what it does with the raw distances next. Two constants
are computed per point:

\begin{definition}[Local calibration]\label{def:rhosigma}
For each point $i$, set $\rho_i = \min_{j \in N(i)} d(x_i, x_j)$ (distance to
the nearest neighbor), and let $\sigma_i > 0$ solve
\begin{equation}\label{eq:sigma}
\sum_{j \in N(i)} \exp\!\left(- \frac{\max(0,\, d(x_i, x_j) - \rho_i)}
{\sigma_i}\right) \;=\; \log_2 k .
\end{equation}
The directed membership strength of edge $i \to j$ is then
\[
w_{i \to j} = \exp\!\left(- \frac{\max(0,\, d(x_i, x_j) - \rho_i)}
{\sigma_i}\right).
\]
\end{definition}

\begin{lemma}[The calibration is well-posed]\label{lem:sigma}
Suppose not all neighbors of $i$ are at distance exactly $\rho_i$ (generic
position). Then Equation~\eqref{eq:sigma} has a unique solution
$\sigma_i \in (0, \infty)$.
\end{lemma}

\begin{proof}
Let $g(\sigma)$ denote the left side, a finite sum of terms
$\exp(-c_j / \sigma)$ with constants $c_j = \max(0, d_{ij} - \rho_i) \ge 0$,
at least one $c_j = 0$ (the nearest neighbor) and, by assumption, at least
one $c_j > 0$. Each term is continuous and nondecreasing in $\sigma$; terms
with $c_j > 0$ are strictly increasing. Hence $g$ is continuous and strictly
increasing, with
\[
\lim_{\sigma \to 0^+} g(\sigma) = \#\{j : c_j = 0\} \ge 1,
\qquad
\lim_{\sigma \to \infty} g(\sigma) = k .
\]
Since $1 \le \log_2 k < k$ for $k \ge 2$ (here $\log_2 15 \approx 3.91$),
the intermediate value theorem gives existence whenever
$\#\{j : c_j = 0\} \le \log_2 k$, and strict monotonicity gives uniqueness.
(The implementation solves \eqref{eq:sigma} by binary search.)
\end{proof}

\begin{remark}[Why calibrate?]\label{rem:uniform}
Equation~\eqref{eq:sigma} forces every point to distribute the same total
membership mass ($\log_2 k$) over its neighborhood. In a dense region,
$\sigma_i$ comes out small (distances are shrunk aggressively); in a sparse
region, large. This is the ``Uniform'' in UMAP: \emph{assume} the data is
uniformly distributed on some manifold, and absorb observed density
variation into a point-dependent rescaling of distance. The $\rho_i$ shift
additionally guarantees every point is connected with weight $1$ to its
nearest neighbor, so no point is orphaned. The cost of this assumption:
UMAP largely erases density information --- a tight cluster and a loose
cluster of the same membership look alike on the map. Cluster \emph{size} on
the galaxy is not evidence of anything.
\end{remark}

Directed weights disagree ($w_{i \to j} \neq w_{j \to i}$), so they are
merged:

\begin{definition}[Fuzzy union]\label{def:fuzzyunion}
The symmetric weight of the pair $\{i, j\}$ is
\[
\mu_{ij} \;=\; w_{i \to j} + w_{j \to i} - w_{i \to j}\, w_{j \to i}.
\]
\end{definition}

\begin{remark}
If the two directed weights are read as independent probabilities that ``$i$
accepts $j$ as a neighbor'' and vice versa, $\mu_{ij}$ is exactly
$\Pr(\text{at least one accepts})$ --- the probabilistic OR
($t$-conorm). One-sided friendships survive; that keeps bridges between
regions alive.
\end{remark}

\section{UMAP II: The Layout}

Act two seeks low-dimensional points $y_1, \dots, y_n \in \R^2$ whose
\emph{own} fuzzy graph resembles $\mu$. The low-dimensional membership is
defined by a smooth rational kernel:

\begin{definition}[Low-dimensional kernel]\label{def:nukernel}
Fix $a, b > 0$. For $y_i, y_j \in \R^2$,
\[
\nu_{ij} \;=\; \Big(1 + a\,\norm{y_i - y_j}^{2b}\Big)^{-1}.
\]
The constants $a, b$ are fit by least squares so that $\nu$ approximates
\[
\Psi(t) =
\begin{cases}
1 & t \le \texttt{min\_dist}, \\
\exp\!\big(-(t - \texttt{min\_dist})\big) & t > \texttt{min\_dist},
\end{cases}
\]
i.e.\ ``full membership within \texttt{min\_dist}, exponential decay
beyond.''
\end{definition}

\texttt{min\_dist} is therefore a pure \emph{rendering} parameter: it sets
how tightly the layout may pack points that the graph considers identical,
trading cluster crispness against within-cluster detail.

\begin{definition}[Objective]\label{def:umapce}
The layout minimizes the fuzzy cross-entropy
\[
\mathcal{C}(y) \;=\; \sum_{i < j}
\underbrace{\mu_{ij} \log \frac{\mu_{ij}}{\nu_{ij}}}_{\text{attraction}}
\;+\;
\underbrace{(1 - \mu_{ij}) \log \frac{1 - \mu_{ij}}{1 - \nu_{ij}}}_{\text{repulsion}} .
\]
\end{definition}

The names are earned by the gradients: the first term is minimized by making
$\nu_{ij}$ large (pull connected points together), the second by making
$\nu_{ij}$ small (push unconnected points apart). The final map is the
equilibrium of $n(n-1)/2$ simultaneous tugs-of-war.

\begin{proposition}[Attractive gradient]\label{prop:grad}
For the attractive term of a single pair, with
$r = \norm{y_i - y_j}$,
\[
\frac{\partial}{\partial y_i}
\Big[ -\mu_{ij} \log \nu_{ij} \Big]
= \mu_{ij}\,
\frac{2ab\, r^{2(b-1)}}{1 + a r^{2b}}\,
\big(y_i - y_j\big),
\]
a spring whose stiffness decays as points approach --- unlike a Hookean
spring, distant neighbors are pulled hard while satisfied ones are left
alone.
\end{proposition}

\begin{proof}
$-\log \nu_{ij} = \log(1 + a r^{2b})$; apply the chain rule with
$\partial r^2 / \partial y_i = 2(y_i - y_j)$ and simplify.
\end{proof}

Minimization proceeds by stochastic gradient descent over \emph{edges}: each
epoch samples edges proportionally to $\mu_{ij}$, applies
Proposition~\ref{prop:grad} to the endpoints, and for each such update also
samples a handful of random non-neighbors ($\approx 5$) to apply the
repulsive gradient to --- \emph{negative sampling}, the standard trick for
avoiding the $\Theta(n^2)$ full repulsion sum. The learning rate anneals
linearly to zero over \texttt{n\_epochs} (the local run used $500$).

Initialization is not random: the points start at the \emph{spectral
embedding} of the graph --- coordinates from the bottom nontrivial
eigenvectors of the symmetric normalized graph Laplacian
$L = I - D^{-1/2} M D^{-1/2}$ (where $M = (\mu_{ij})$ and $D$ is its degree
matrix). This is exactly the Laplacian-eigenmaps solution, a convex
relaxation of the same ``neighbors stay close'' demand, and it starts SGD in
a basin that already respects the coarse cluster structure.

\section{What the Map Does and Does Not Mean}

\begin{proposition}[Isometry invariance]\label{prop:invariance}
Let $T : \R^2 \to \R^2$ be any Euclidean isometry (rotation, reflection,
translation, or a composition). Then
$\mathcal{C}(T y_1, \dots, T y_n) = \mathcal{C}(y_1, \dots, y_n)$.
\end{proposition}

\begin{proof}
$\mathcal{C}$ depends on the layout only through the pairwise distances
$\norm{y_i - y_j}$ (via $\nu_{ij}$), and isometries preserve pairwise
distances by definition.
\end{proof}

\begin{corollary}[The axes are meaningless]\label{cor:axes}
The set of minimizers of $\mathcal{C}$ is closed under all Euclidean
isometries. Which representative the algorithm returns is decided by the
spectral initialization and the randomness of SGD --- not by the data. In
particular, no direction on the galaxy map (up, down, left, right) admits
any interpretation, and separate runs on the same library may return maps
differing by rotation, reflection, or more.
\end{corollary}

What survives, and why one may trust it: $\mu_{ij}$ is built \emph{only}
from each point's $k$ nearest neighbors, and the objective's attraction acts
only on those pairs. Local neighborhoods are the signal; everything else ---
inter-cluster distances, cluster areas (Remark~\ref{rem:uniform}), global
shape --- is a side effect of repulsion and should be read with suspicion.

\begin{incode}
\texttt{compute\_galaxy.py} runs
\texttt{umap.UMAP(n\_components=2, n\_neighbors=15, min\_dist=0.05,
metric="cosine")} for positions. Post-processing: positions are centered and
divided by the $99.5$th percentile of coordinate magnitude --- a robust
rescale to roughly $[-1,1]^2$ that one outlier moment cannot hijack.
The client (\texttt{app.js}, \texttt{Galaxy}) and the drift view
(\texttt{vibescenes.js}) treat the coordinates as opaque: everything they
draw --- hover neighborhoods, the song path, the flight camera --- uses only
relative position, as Corollary~\ref{cor:axes} demands.
\end{incode}

\section{Color as a Second Opinion}

The star colors are not styling; they are a second, independent instance of
the same mathematics. The pipeline runs UMAP \emph{again} on the same
$64$-dimensional data with \texttt{n\_components=3} and a looser
\texttt{min\_dist} of $0.3$, yielding $z_i \in \R^3$, then maps robustly to
RGB:
\[
\text{rgb}_i \;=\; 60 + 195 \cdot
\operatorname{clip}\!\left(
\frac{z_i - p_2}{p_{98} - p_2},\; 0,\; 1\right),
\]
where $p_2, p_{98}$ are per-channel percentiles (the affine map lands in
$[60, 255]^3$ so no star is too dark against the background).

Because the 3-D run shares the input graph-building mathematics but not the
2-D run's optimization trajectory, \emph{hue similarity is a
quasi-independent witness of timbral similarity}. When two nearby stars also
share a color, two different embeddings agreed; when a star's color clashes
with its neighborhood, the 2-D layout likely forced a compromise
(Corollary~\ref{cor:axes} again: the plane is small).

\section*{Exercises}

\begin{exercise}
Prove the identity $\norm{u - v}^2 = 2\,d_{\cos}(u,v)$ for unit vectors, and
exhibit three unit vectors in $\R^2$ violating the triangle inequality for
$d_{\cos}$.
\end{exercise}

\begin{exercise}
Prove the subspace half of Theorem~\ref{thm:pca}: among all $k$-dimensional
subspaces $S$, total projected variance is maximized by the span of the top
$k$ eigenvectors. (Induct using the variational characterization on the
orthogonal complement.)
\end{exercise}

\begin{exercise}
In Lemma~\ref{lem:sigma}, suppose point $i$ has $t$ neighbors tied exactly at
distance $\rho_i$ with $t > \log_2 k$. What goes wrong, and what does the
condition mean geometrically? (The implementation clamps $\sigma_i$ to a
minimum value in this case.)
\end{exercise}

\begin{exercise}
Verify Proposition~\ref{prop:grad}. Then derive the repulsive gradient
$\partial\big[-(1-\mu_{ij})\log(1 - \nu_{ij})\big] / \partial y_i$ and show
it diverges as $r \to 0$ when $b < 1$. The implementation adds a small
$\varepsilon$ to $r^2$; explain what failure this prevents during SGD.
\end{exercise}

\begin{exercise}
Prove that $\mu_{ij} = w_{i\to j} + w_{j \to i} - w_{i \to j} w_{j \to i}$
satisfies $\max(w_{i \to j}, w_{j \to i}) \le \mu_{ij} \le
\min(1,\, w_{i \to j} + w_{j \to i})$, and characterize when each bound is
attained.
\end{exercise}

\begin{exercise}[computational]
Run \texttt{compute\_galaxy.py} twice with different random seeds and overlay
the two maps. Find the best-fitting rotation+reflection between them by
solving the orthogonal Procrustes problem
$\min_{Q^\T Q = I} \norm{Y_1 Q - Y_2}_F$ (solution: $Q = UV^\T$ from the SVD
of $Y_1^\T Y_2$). How much residual disagreement survives after aligning?
Which regions of the map are stable, and are they the dense ones?
\end{exercise}

\begin{exercise}[computational]
Set \texttt{n\_neighbors=5} and rebuild the galaxy. Describe what happens to
(i) song threads, (ii) the number of visual clusters, (iii) bridges between
genres. Explain each observation via Remark~\ref{rem:uniform} and the role of
$k$ in Definition~\ref{def:knn}.
\end{exercise}
